---
title: "Homework Assignment #8"
author: "Maria Kunevich"
date: "2022-12-01"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Assignment Description  

Download the dataset from the website on Estonian statistics into `R Markdown` session. The number of admitted students across different specialties is recorded for Year 2021. The numbers of admitted students are presented for *Study programme group*, *Level of study*, and *Mother tongue*.     

To download the dataset, use the following R code:  
```{r, include=FALSE}
library(readr)
library(dplyr)

# download the dataset from the Github repository using this code

myfile <- "https://raw.githubusercontent.com/Maria-13/DataAnalytics_R/main/Assignments/higher_edu_in_Estonia-1.csv"
higher.edu <- read.csv(url(myfile))
View(higher.edu)
higher.edu <- as_tibble(higher.edu)
# delete the second column 'Indicator'
higher.edu <- select(higher.edu, -Year, -Indicator)
```
1. Perform quick EDA and pick up variables you want to explore in more depth.  

2. Prepare the data set that includes only variables of your interest in a suitable format for analysis (use `dlyr` package and `tidyr` when necessary).  

3. Based on the dataset, formulate your *Research Question(s)*. Think of questions that you can answer with the help of either *chi-square* statistical test or some type of *t-test*. 

4. Hypothesis testing.   
4.1 State the null hypothesis and the alternative hypothesis.   
4.2 Report on collected data and sample size.  
4.3 Check the assumption of the chosen statistical test. Perform the required statistical test.  
4.4 Decide whether to reject or fail to reject your null hypothesis, report selected significance level.  
4.5 Interpret and report the results.  

5. Create a report in `R Markdown` with clear sections, using headings and subheadings.  





